### 2017-12-07 paper/wind.tex
All experiments performed

\documentclass[]{article}

\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{float}
\usepackage{hyperref}

\title{Wind Prediction with LSTM}
\author{Juan Lao Tebar}

\begin{document}

\maketitle

\begin{abstract}

	TODO

\end{abstract}

\section{Introduction}

In this practice we extend the experiments performed in the previous ``Guided Laboratory of the RNN Topic''\footnote{https://upc-mai-dl.github.io/rnn-lab-guided/}, not only predicting more steps and trying more complex topologies, but also testing different ways of preparing the data to exploit the true power of LSTM networks, learning the inherent data relations over long intervals of time.

As explained in the guided laboratory, the data for this example is a subset of the NREL Wind Integration National Dataset (WIND) Toolkit\footnote{https://www.nrel.gov/grid/wind-toolkit.html}. The original set consists of meteorological information for more than 126,000 sites across the USA from 2007 to 2013, sampled every 5 minutes. However, for our purposes, we only take into account the data from one site sampled every 15 minutes.

In the guided laboratory we took into account only one variable---wind speed at 100m---and we predicted the value it takes in the following step. In contrast, in this practice we predict the following 6 steps including the rest of the variables: air density, temperature and air pressure.

The workflow of this practice consists in defining a baseline model---similar to the one proposed in the guided laboratory---and improving it in the following experiments, step by step.

In summary, these are the experiments we perform:

\begin{itemize}
	\item Effect of incrementing the number of hidden layers.
	\item Response of the model for different activation functions.
	\item Finding the appropriate window size.
	\item Regularizing an overfitted model applying dropout on \emph{all} activations---including LSTM gates.
	\item Avoiding explosive gradients with gradient clipping.
	\item Regularizing an overfitted model applying dropout \emph{only} on inter-layer connections.
	\item Presenting a new sequence-to-sequence model capable of learning from \emph{any-length} input samples.
\end{itemize}

The code related to this work is public and available at github\footnote{https://github.com/juanlao7/WIND-LSTM}.

\section{Experiments}

The original dataset consists of 245,472 measurements. In our process we use 122,736 of them for training---\~3.5 years---, 61,368 for validation and 61,368 for testing---\~1.75 years---. We decided to train and test our model in these large intervals of time due to the fact that weather depends on the time of the year and, since LSTM networks can learn data relations over long intervals of time, we may obtain better results when testing huge window sizes.

The loss of the model is measured with the mean squared error (MSE).

\subsection{Baseline}

Our first model is quite similar to the one proposed in the previous guided laboratory, as it consists of an LSTM layer of 512 units connected to a dense output layer of 24 units---6 predictions of 4 variables---, as it can be seen in figure \ref{f:model}. We use ReLU as the activation function of the LSTM activations, the logistic function in their recurrent step and, of course, a linear function for the output neurons.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{model}
	\caption{Structure of the baseline model.}
	\label{f:model}
\end{figure}

% TODO: citar rmsprop
% TODO: buscar ref de recommended rmsprop

We train the model in batches of 1000 samples using the adaptive optimizer RMSprop---recommended for training recurrent neural networks---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay. The samples are obtained by \emph{lagging} the original samples of the data set with their previous 6 steps and foreseeing the following 6.

In this practice we change the stopping criteria of the training process, as we remove the limit of epochs and we implement a stopper based on the validation loss: when we detect a minimum value, we stop the training after 30 epochs if it does not decrease.

With this configuration we obtain the results shown in table \ref{t:baseline}.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ccc@{}}
		\toprule
		Training loss & Validation loss & Test loss \\ \midrule
		0.02878       & 0.03151         & 0.03290   \\ \bottomrule
	\end{tabular}
	\caption{Results of the baseline model.}
	\label{t:baseline}
\end{table}

\subsection{Number of Layers}

% TODO ref esto

Feedforward neural networks with two hidden layers can represent an arbitrary decision boundary with arbitrary accuracy using rational activation functions, and approximate any smooth variable relation to any accuracy too.

In this experiment we compare our previous results with the ones obtained with models of two and three hidden LSTM layers of 512 units each.

Table \ref{t:layers} shows the results obtained and figure \ref{f:layers} shows the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}cccc@{}}
		\toprule
		& Training loss & Validation loss & Test loss \\ \midrule
		1 hidden layer  & 0.02878       & 0.03151         & 0.03290   \\
		2 hidden layers & 0.02771       & 0.03088         & 0.03257   \\
		3 hidden layers & 0.02830       & 0.03207         & 0.03327   \\ \bottomrule
	\end{tabular}
	\caption{TODO}
	\label{t:layers}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{layers}
	\caption{TODO}
	\label{f:layers}
\end{figure}

As expected, 2 hidden layers approximate better our data, while 3 hidden layers does not seem to have any effect. In the next experiments we use an architecture of 2 hidden LSTM layers.

\subsection{Activation Function}

Clevert, Unterthiner, \& Hochreiter, 2016 \cite{clevert2015fast} found that using ELU instead of ReLU not only speeds up the learning process in deep neural networks, but also leads to higher classification accuracies: ``In contrast to ReLUs, ELUs have negative values which allows them to push mean unit activations closer to zero like batch normalization but with lower computational complexity. Mean shifts toward zero speed up learning by bringing the normal gradient closer to the unit natural gradient because of a reduced bias shift effect''.

In this experiment we aim to reproduce similar results, testing different activation functions---softplus, logistic, hyperbolic tangent, ELU and ReLU---for the hidden LSTM layers of our model, but still using the logistic function for the recurrent step.

Table \ref{t:activation} shows the results obtained, while figure \ref{f:activation} shows the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ccccc@{}}
		\toprule
		& Training loss & Validation loss & Test loss & Epochs \\ \midrule
		Softplus           & 0.02972       & 0.03352         & 0.03485   & 288    \\
		Logistic           & 0.02986       & 0.03217         & 0.03348   & 622    \\
		Hyperbolic tangent & 0.02864       & 0.03133         & 0.03289   & 172    \\
		ELU                & 0.02856       & 0.03143         & 0.03251   & 186    \\
		ReLU               & 0.02771       & 0.03088         & 0.03257   & 206    \\ \bottomrule
	\end{tabular}
	\caption{TODO}
	\label{activation}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{activation}
	\caption{TODO}
	\label{f:activation}
\end{figure}

As we can see, in our case ELUs and ReLUs are quite similar. We can see how ELUs take a slightly lower number of epochs to converge, with almost the same test loss then with ReLUs but higher training and validation losses.

In the following experiments we keep ReLU as the activation function for all the units of the hidden layers.

\subsection{Window Size}

Our data set is generated by going through the original samples with a window size of 12---6 measurements for the input, 6 for the output---. We must note that these measurements have been taken in intervals of 15 minutes, hence each input vector of the generated dataset comprises an interval of 75 minutes.

In this experiment we test different window sizes, in order to determine if larger time intervals provide useful information for our model to obtain a lower loss.

Table \ref{t:window} shows the results obtained, while figure \ref{f:window} shows the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}cccc@{}}
		\toprule
		& Training loss & Validation loss & Test loss \\ \midrule
		Input size of 1  & 0.03595       & 0.03743         & 0.03810   \\
		Input size of 6  & 0.02771       & 0.03088         & 0.03257   \\
		Input size of 12 & 0.02655       & 0.03190         & 0.03345   \\
		Input size of 24 & 0.02242       & 0.03655         & 0.03743   \\
		Input size of 48 & 0.01975       & 0.04247         & 0.04327   \\ \bottomrule
	\end{tabular}
	\caption{TODO}
	\label{window}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{window}
	\caption{TODO}
	\label{f:window}
\end{figure}

In this experiment we test window sizes up to 48. When we test higher values, at some point the loss becomes $ +\inf $ and the training process fails. There can be several reasons behind this behaviour, but we suspect that it is due to a combination of using ELUs---which range is $ [-1, +\inf] $---and a high learning rate. However, when we try to use the logistic function we get worse results---as found in the previous experiment---and, when we decrease the learning rate, our experiments take too long to finish. For these reasons we decide to ignore bigger window sizes.

Regarding to the results, as we can see, bigger windows lead to lower training losses. In fact, our model overfits the data when we train it with sequences of 24 or 48 inputs, increasing the validation loss while decreasing the training error without any sign of convergence.

In the following experiments we continue working with the model with the best test loss---trained with sequences of 6 inputs---and the model with the best training loss---48 inputs.

\subsection{Dropout I}

Dropout is considered a simple and powerful regularization method. \cite{srivastava2014dropout} ``The key idea is to randomly drop units (along with their connections) from the neural network during training. This prevents units from co-adapting too much. During training, dropout samples from an exponential number of different ``thinned'' networks. At test time, it is easy to approximate the effect of averaging the predictions of all these thinned networks by simply using a single unthinned network that has smaller weights. This significantly reduces overfitting and gives major improvements over other regularization methods''.

In this experiment we apply different dropout values---25\%, 50\% and 75\%---in each LSTM layer, not only at the inter-layer connections but also at the recurrent step.

Table \ref{t:drop1} shows the results obtained, while figures \ref{f:drop11} and \ref{f:drop12} show the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ccccc@{}}
		\toprule
		&      & Training loss             & Validation loss & Test loss       \\ \midrule
		Input size of 6  & 0\%  & 0.02771                   & 0.03088         & 0.03257         \\
		& 25\% & 0.14523                   & 0.09438         & 0.09005         \\
		& 50\% & 0.32983                   & 0.27392         & 0.25589         \\
		& 75\% & 0.63196                   & 0.55140         & 0.50822         \\
		\midrule
		Input size of 48 & 0\%  & 0.02231                   & 0.03864         & 0.04073         \\
		& 25\% & 40,982,524.64120          & 2,030,493.06455 & 2,327,123.48740 \\
		& 50\% & 0.29737                   & 0.26379         & 0.24811         \\
		& 75\% & 395,288,170,387,000.00000 & 191,750.24100   & 210,455.33092   \\ \bottomrule
	\end{tabular}
	\caption{TODO}
	\label{t:drop1}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{drop11}
	\caption{TODO}
	\label{f:drop11}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{drop12}
	\caption{TODO}
	\label{f:drop12}
\end{figure}

% TODO ref exploding gradients
As we can see, on input sizes of 48 samples, applying dropout cause a problem of exploding gradients, and our losses take very large values.

% TODO gradient clipping
Gradient clipping prevents gradients from exploding. As we can see in table \ref{t:drop2} and figure \ref{f:drop2}, when all parameter gradients are clipped to a maximum norm of 1 we can successfully train our models with dropout.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ccccc@{}}
		\toprule
		&      & Training loss & Validation loss & Test loss \\ \midrule
		Input size of 48 & 0\%  & 0.02248       & 0.03962         & 0.03925   \\
		& 25\% & 0.11694       & 0.09555         & 0.09036   \\
		& 50\% & 0.28648       & 0.27258         & 0.25334   \\
		& 75\% & 0.59369       & 0.51817         & 0.48055   \\ \bottomrule
	\end{tabular}
	\caption{My caption}
	\label{t:drop2}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{drop2}
	\caption{TODO}
	\label{f:drop2}
\end{figure}

In our case, gradient clipping has demonstrated to be a good method for avoiding exploding gradients, hence we will conserve it in the following experiments. However, applying dropout at the inter-layer connections and at the recurrent step does not improve the performance of our model at all; on the contrary, it gets much worse.

\subsection{Dropout II}

Taking into account that LSTMs solve the vanishing gradient problem using an internal state controlled by gates, one may think that it may not be a good idea to apply dropout over those gates. It is possible that the bad results obtained in the previous experiment can be prevented just applying dropout at the inter-layer connections but not at the recurrent step.

In this experiment we apply different dropout values---25\%, 50\% and 75\%---after each LSTM layer.

Table \ref{t:drop3} shows the results obtained, while figures \ref{f:drop31} and \ref{f:drop32} show the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ccccc@{}}
		\toprule
		&      & Training loss & Validation loss & Test loss \\ \midrule
		Input size of 6  & 0\%  & 0.02707       & 0.03102         & 0.03236   \\
		& 25\% & 0.03272       & 0.03107         & 0.03227   \\
		& 50\% & 0.04334       & 0.03215         & 0.03352   \\
		& 75\% & 0.06819       & 0.03617         & 0.03711   \\
		Input size of 48 & 0\%  & 0.02231       & 0.03864         & 0.04073   \\
		& 25\% & 0.03074       & 0.03451         & 0.03576   \\
		& 50\% & 0.03949       & 0.03425         & 0.03584   \\
		& 75\% & 0.06712       & 0.03815         & 0.03869   \\ \bottomrule
	\end{tabular}
	\caption{My caption}
	\label{my-label}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{drop31}
	\caption{TODO}
	\label{f:drop31}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{drop32}
	\caption{TODO}
	\label{f:drop32}
\end{figure}

As we can see, dropout does make any considerable improvement on the case in which the input size is 6---probably because that model was not overfitting the training data---; however, when the input size is 48, we can see how a dropout of 25\% successfully reduces the gap between training and validation losses, obtaining an overall lower test error.

In any case, the model with the bigger window size still has a worse performance than the model with an input size of 6. In the following experiment we only test this last model without dropout.

\subsection{Sequential Architecture}

LSTM networks can learn the inherent data relations between over intervals of time of \emph{any} length; however, in our current architecture, we are feeding the model with short subsequences generated with a small sliding window.

Why are we doing that? Can we train our LSTM with all our data set as a single sequence? If we do that, the LSTM network will learn whatever it finds relevant. It may happen that it finds data relations over large intervals of time, improving the overall performance of the system.

In our current architecture, if we try to process our data set with a huge window size, we find a physical problem.

When we process the data set, we go through $ N $ samples of $ V $ variables with a window size of $ W $, generating $ N - W $ vectors of $ V \cdot (W + 1) $ variables each. Since we train in batches of 1000, we are working with $ 1000 \cdot V \cdot (W + 1) $ variables simultaneously.

The memory used by this architecture directly depends on the size of sequences we use for training, making it not scalable for training with very large sequences. We would need large amounts of RAM to exploit the true power of LSTM networks.

To overcome this obstacle, in this experiment we present a new architecture with a model that gives a prediction for each input we feed to it (i.e. when we feed a sequence of $ N $ inputs, the model gives $ N $ outputs). This new architecture completely discards the sliding window process and directly feeds the original measurements to the model, targeting the next 6 measurements of the data set for each given input.

This new model uses an option defined by Keras framework in their LSTM layer implementation\footnote{https://keras.io/layers/recurrent/\#lstm}, the argument ``stateful'': ``Boolean (default False). If True, the last state for each sample at index $ i $ in a batch will be used as initial state for the sample of index $ i $ in the following batch''.

In this new architecture each element of the training sequences pertain to different batches and we feed them one by one, hence the RAM used by the training process is constant and it does not depend on the sequences size. In fact, now the required RAM only depends on the batch size we choose.

Our data is not compatible with this new training process and we must make a conversion first, since it is sequentially sorted by time. This conversion consists in reordering our data in such a way that, for a data set of $ N $ samples and a batch size of $ B $, samples from $ s_1 $ to $ s_{\frac{N}{B}} $ end up being the first elements of every batch, in sequential order; then samples from $ s_{\frac{N}{B}+1} $ to $ s_{\frac{N}{B}+\frac{N}{B}} $ end up being the second elements of every batch and so on. See figure \ref{f:explanation} for more details.

TODO EXPLANATION

As we can see, with this conversion we are training our model with $ B $ different sequences. Here we can decide how long we want our sequences: with a batch size of 1 we will train our model with a single sequence of 122,736 samples; with a batch size of 4 we will train our model 4 sequences of size 30,684.

In this experiment we leave the batch size as 1000---we train with input sequences of 122 samples---, in order to compare the results with the ones obtained in the previous experiments. We must leave the same batch size because it directly affects to the model’s generalization.

In this experiment we also clip all parameter gradients to a maximum norm of 0.000000001. Unfortunately we still suffer from exploding gradients, as it can be seen in figure \ref{f:architecture1}. This strict clipping slows the learning process; note that the gradients explode after after 2964 epochs---1 hour and 39 minutes---and the training loss is not even below 0.8. We could test lower clipping norms but then the training process would take an unfeasible amount of time to finish.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{architecture1}
	\caption{TODO}
	\label{f:architecture1}
\end{figure}

Our gradients are exploding due to the usage of ReLU, since its range is $ [0, +\inf] $. If we replace it with the hyperbolic tangent function---range $ [-1, 1] $---then we do not suffer from this problem.

Table \ref{t:architecture2} shows the results obtained, while figure \ref{f:architecture2} shows the evolution of the training and validation loss in each case.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}cccc@{}}
		\toprule
		& Training loss & Validation loss & Test loss \\ \midrule
		Sequential      & 0.03556       & 0.04878         & 0.05231   \\
		Input size of 6 & 0.02707       & 0.03102         & 0.03236   \\ \bottomrule
	\end{tabular}
	\caption{My caption}
	\label{architecture2}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\textwidth]{architecture2}
	\caption{TODO}
	\label{f:architecture2}
\end{figure}

This new architecture may be useful for problems where the model needs to recall information far from the past (e.g., stock price prediction or paragraph understanding) but it does not offer any advantage over our previous architecture based on a sliding window. From the previous experiments, we already had the suspicion that all the relevant information for making a good prediction relies on the short past.

\section{Discussion of the Results}

\section{Conclusion and Future Work}

\end{document}

### 2017-12-07 baseline.py

@@ -1,181 +1,0 @@
-"""
-.. module:: WindPrediction
-
-WindPrediction
-*************
-
-:Description: WindPrediction
-
-:Authors: bejar, lao
-    
-
-:Version: 
-
-:Created on: 06/09/2017 9:47
-
-:Updated on: 22/11/2017 0:20 
-
-"""
-
-import numpy as np
-
-from keras.models import Sequential
-from keras.layers import Dense, Activation
-from keras.layers import LSTM, GRU
-from keras.optimizers import RMSprop, SGD
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import StandardScaler
-from sklearn.metrics import mean_squared_error
-import argparse
-import json
-import time
-from keras.callbacks import EarlyStopping
-
-__author__ = 'bejar'
-
-def lagged_vector(data, lag=1):
-    """
-    Returns a matrix with columns that are the steps of the lagged time series
-    Last columns are the value to predict
-    :param data:
-    :param lag:
-    :return:
-    """
-    lvect = []
-    for i in xrange(lag):
-        lvect.append(data[i: -lag+i])
-    lvect.append(data[lag:])
-    lvect = np.stack(lvect, axis=1)
-    return np.reshape(lvect, (lvect.shape[0], lvect.shape[1] * lvect.shape[2]))
-
-def load_config_file(nfile, abspath=False):
-    """
-    Read the configuration from a json file
-
-    :param abspath:
-    :param nfile:
-    :return:
-    """
-    ext = '.json' if 'json' not in nfile else ''
-    pre = '' if abspath else './'
-    fp = open(pre + nfile + ext, 'r')
-
-    s = ''
-
-    for l in fp:
-        s += l
-
-    return json.loads(s)
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser()
-    parser.add_argument('--config', default='config', help='Experiment configuration')
-    parser.add_argument('--verbose', help="Verbose output (enables Keras verbose output)", action='store_true', default=False)
-    parser.add_argument('--gpu', help="Use LSTM/GRU gru implementation", action='store_true', default=False)
-    args = parser.parse_args()
-
-    verbose = 1 if args.verbose else 0
-    impl = 2 if args.gpu else 0
-
-    config = load_config_file(args.config)
-
-    print("Starting:", time.ctime())
-
-    ###########################################
-    # Data
-
[…]

### 2017-12-07 process.py

@@ -1,0 +1,291 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47
+
+:Updated on: 22/11/2017 0:20 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation, Dropout
+from keras.layers import LSTM, GRU
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+
+__author__ = 'bejar, lao'
+
+FUTURE = 6
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    plt.ylim(ymin=0.02, ymax=0.06)
+    
+    plt.show()
+    
+# Makes the dataset compatible with keras stateful LSTM's, where the ith element of the batch uses the state stored for the ith element of the previous batch.
+def batchify((x, y), batchSize):
+    x, y = removeSurplus((x, y), batchSize)
+    x = batchifyImpl(x, batchSize)
+    y = batchifyImpl(y, batchSize)
+    return x, y
+
+def batchifyImpl(data, batchSize):
+    n = len(data)
+    parsedData = np.empty(data.shape)
+    k = 0
+    
+    for i in xrange(batchSize):
+        for j in xrange(i, n, batchSize):
+            parsedData[k] = data[j]
+            k += 1
+    
+    return parsedData
+
+def removeSurplus((x, y), batchSize):
+    surplus = x.shape[0] % batchSize
+    
+    if surplus == 0:
[…]

### 2017-12-07 paper/wind.tex
Finished, waiting for review

@@
 
-	TODO
+LSTM neural networks are powerful learning models, specially on long sequences with relations in long intervals of time.
+In this practice we analyze their behavior against the WIND dataset, implementing several Deep Learning techniques to improve the learning process and to prevent overfitting.
 
@@
 
-In summary, these are the experiments we perform:
+More in detail, the plan consists in the following:
 
-\begin{itemize}
-	\item Effect of incrementing the number of hidden layers.
-	\item Response of the model for different activation functions.
-	\item Finding the appropriate window size.
-	\item Regularizing an overfitted model applying dropout on \emph{all} activations---including LSTM gates.
-	\item Avoiding explosive gradients with gradient clipping.
-	\item Regularizing an overfitted model applying dropout \emph{only} on inter-layer connections.
-	\item Presenting a new sequence-to-sequence model capable of learning from \emph{any-length} input samples.
-\end{itemize}
+\begin{enumerate}
+	\item Study the effect of incrementing the number of hidden layers.
+	\item Observe the response of the model for different activation functions.
+	\item Find a better window size.
+	\item If the model overfits the training data, apply dropout---on inter-layer connections and/or on LSTM gates.
+	\item In case of experimenting explosive gradients, avoid them with gradient clipping.
+	\item Present a new sequence-to-sequence model capable of learning from \emph{any-length} input samples.
+\end{enumerate}
 
@@
 
-The original dataset consists of 245,472 measurements.
-In our process we use 122,736 of them for training---\~3.5 years---, 61,368 for validation and 61,368 for testing---\~1.75 years---.
-We decided to train and test our model in these large intervals of time due to the fact that weather depends on the time of the year and, since LSTM networks can learn data relations over long intervals of time, we may obtain better results when testing huge window sizes.
+The original data set consists of 245,472 measurements. In our process we use 122,736 of them for training---$\sim$3.5 years---, 61,368 for validation and 61,368 for testing---$\sim$1.75 years.
 
@@
 Our first model is quite similar to the one proposed in the previous guided laboratory, as it consists of an LSTM layer of 512 units connected to a dense output layer of 24 units---6 predictions of 4 variables---, as it can be seen in figure \ref{f:model}.
-We use ReLU as the activation function of the LSTM activations, the logistic function in their recurrent step and, of course, a linear function for the output neurons.
+We use ReLU as the activation function of the LSTM units, the logistic function for their recurrent step and, of course, a linear function for the output neurons.
 
@@
 
-% TODO: citar rmsprop
-% TODO: buscar ref de recommended rmsprop
-
-We train the model in batches of 1000 samples using the adaptive optimizer RMSprop---recommended for training recurrent neural networks---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay.
+We train the model in batches of 1000 samples using the adaptive optimizer RMSprop \cite{tieleman2012lecture}---recommended for training recurrent neural networks---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay.
 The samples are obtained by \emph{lagging} the original samples of the data set with their previous 6 steps and foreseeing the following 6.
@@
 
-% TODO ref esto
+Feedforward neural networks with two hidden layers can represent any decision boundary with arbitrary accuracy using the proper number of units, as well as approximating any smooth variable relation to any accuracy too.
 
-Feedforward neural networks with two hidden layers can represent an arbitrary decision boundary with arbitrary accuracy using rational activation functions, and approximate any smooth variable relation to any accuracy too.
+For this reason, in this experiment we compare our previous results with the ones obtained with models of two and three hidden LSTM layers of 512 units each.
 
-In this experiment we compare our previous results with the ones obtained with models of two and three hidden LSTM layers of 512 units each.
-
-Table \ref{t:layers} shows the results obtained and figure \ref{f:layers} shows the evolution of the training and validation loss in each case.
+Table \ref{t:layers} shows the results obtained, while figure \ref{f:layers} shows the evolution of the training and validation loss in each case.
 
@@
 	\end{tabular}
-	\caption{TODO}
+	\caption{Results after testing different number of hidden layers.}
 	\label{t:layers}
@@
 	\includegraphics[width=0.5\textwidth]{layers}
-	\caption{TODO}
+	\caption{Validation and training loss with each number of hidden layers.}
 	\label{f:layers}
@@
 
-As expected, 2 hidden layers approximate better our data, while 3 hidden layers does not seem to have any effect. In the next experiments we use an architecture of 2 hidden LSTM layers.
+As expected, two hidden layers approximate better our data relations, while three hidden layers does not seem to improve the mark.
+In the following experiments we use an architecture of two hidden LSTM layers.
 
@@
 
-In this experiment we aim to reproduce similar results, testing different activation functions---softplus, logistic, hyperbolic tangent, ELU and ReLU---for the hidden LSTM layers of our model, but still using the logistic function for the recurrent step.
+In this experiment we aim to reproduce similar results, testing different activation functions---softplus, logistic, hyperbolic tangent, ELU and ReLU---for the hidden LSTM layers of our model, but still using the logistic function for their recurrent step.
 
@@
 	\end{tabular}
-	\caption{TODO}
-	\label{activation}
+	\caption{Results after testing different activation functions.}
+	\label{t:activation}
 \end{table}
@@
 	\includegraphics[width=\textwidth]{activation}
-	\caption{TODO}
+	\caption{Validation and training loss with each activation function.}
 	\label{f:activation}
@@
 As we can see, in our case ELUs and ReLUs are quite similar.
-We can see how ELUs take a slightly lower number of epochs to converge, with almost the same test loss then with ReLUs but higher training and validation losses.
+We can see how ELUs take a slightly lesser number of epochs to converge, with almost the same test loss as with ReLUs but with higher training and validation losses.
 
-In the following experiments we keep ReLU as the activation function for all the units of the hidden layers.
+In the following experiments we keep ReLU as the activation function for all the units of the hidden layers, since it is computationally cheap and we have not seen any considerable improvement on using ELU.
 
@@
 
-Our data set is generated by going through the original samples with a window size of 12---6 measurements for the input, 6 for the output---.
-We must note that these measurements have been taken in intervals of 15 minutes, hence each input vector of the generated dataset comprises an interval of 75 minutes.
+Our data set is generated by going through the original samples with a window size of 12---6 measurements for the input + 6 for the output---.
+We must note that these measurements have been taken in intervals of 15 minutes, hence each input vector of the generated data set comprises an interval of 75 minutes.
 
@@
 	\end{tabular}
-	\caption{TODO}
-	\label{window}
+	\caption{Results after testing different window sizes.}
+	\label{t:window}
 \end{table}
@@
 	\includegraphics[width=\textwidth]{window}
-	\caption{TODO}
+	\caption{Validation and training loss with each window size.}
 	\label{f:window}
@@
 
-In this experiment we test window sizes up to 48.
-When we test higher values, at some point the loss becomes $ +\inf $ and the training process fails.
-There can be several reasons behind this behaviour, but we suspect that it is due to a combination of using ELUs---which range is $ [-1, +\inf] $---and a high learning rate.
-However, when we try to use the logistic function we get worse results---as found in the previous experiment---and, when we decrease the learning rate, our experiments take too long to finish.
-For these reasons we decide to ignore bigger window sizes.
+In this experiment we test input sizes up to 48.
+When we test higher values, at some point our gradients explode, loss becomes $ +\infty $ and the training process fails.
+There can be several reasons behind this behavior; we suspect that it is due to a combination of using ReLUs---which range is $ [0, +\infty) $---and a high learning rate.
+This problem can be easily solved just applying gradient clipping, but at this point we prefer just not to try bigger window sizes.
+Later, in this document, we will apply gradient clipping in another scenario.
 
 Regarding to the results, as we can see, bigger windows lead to lower training losses.
-In fact, our model overfits the data when we train it with sequences of 24 or 48 inputs, increasing the validation loss while decreasing the training error without any sign of convergence.
+In fact, our model overfits the data when we train it with sequences of 24 or 48 inputs, as we can see how the validation loss increases while training loss decreases without any sign of convergence.
 
-In the following experiments we continue working with the model with the best test loss---trained with sequences of 6 inputs---and the model with the best training loss---48 inputs.
+In the following experiments we continue working with the model trained with sequences of 6 inputs---the best test loss---and the model trained with sequences of 48 inputs---the best training loss.
 
@@
 
-In this experiment we apply different dropout values---25\%, 50\% and 75\%---in each LSTM layer, not only at the inter-layer connections but also at the recurrent step.
+In this experiment we apply different dropout factors---25\%, 50\% and 75\%---in each LSTM layer, not only at the inter-layer connections but also at the recurrent step.
 
@@
 	\end{tabular}
-	\caption{TODO}
+	\caption{Results after testing different dropout factors.}
 	\label{t:drop1}
@@
 	\includegraphics[width=0.5\textwidth]{drop11}
-	\caption{TODO}
+	\caption{Validation and training loss with each dropout factor, for input sizes of 6 samples.}
 	\label{f:drop11}
@@
 	\includegraphics[width=0.5\textwidth]{drop12}
-	\caption{TODO}
+	\caption{Validation and training loss with each dropout factor, for input sizes of 48 samples.}
 	\label{f:drop12}
@@
 
-% TODO ref exploding gradients
-As we can see, on input sizes of 48 samples, applying dropout cause a problem of exploding gradients, and our losses take very large values.
+As we can see, on input sizes of 48 samples, applying a dropout factor makes gradients explode and loss takes very large values.
 
-% TODO gradient clipping
 Gradient clipping prevents gradients from exploding.
-As we can see in table \ref{t:drop2} and figure \ref{f:drop2}, when all parameter gradients are clipped to a maximum norm of 1 we can successfully train our models with dropout.
+As we can see in table \ref{t:drop2} and figure \ref{f:drop2}, when all parameter gradients are clipped to a maximum norm of 1, we can successfully train our models while applying dropout.
 
@@
 	\end{tabular}
-	\caption{My caption}
+	\caption{Results after testing different dropout factors and applying gradient clipping.}
 	\label{t:drop2}
@@
 	\includegraphics[width=0.5\textwidth]{drop2}
-	\caption{TODO}
+	\caption{Validation and training loss with each dropout factor, for input sizes of 48 samples, applying gradient clipping.}
 	\label{f:drop2}
@@
 
-In our case, gradient clipping has demonstrated to be a good method for avoiding exploding gradients, hence we will conserve it in the following experiments.
-However, applying dropout at the inter-layer connections and at the recurrent step does not improve the performance of our model at all; on the contrary, it gets much worse.
+Gradient clipping has demonstrated to be a good method for avoiding exploding gradients, hence we will conserve it in the following experiments.
+On the other hand, we found that applying dropout at the inter-layer connections and at the recurrent step does not improve the performance of our model at all; on the contrary, it gets much worse.
 
@@
 
-Taking into account that LSTMs solve the vanishing gradient problem using an internal state controlled by gates, one may think that it may not be a good idea to apply dropout over those gates.
-It is possible that the bad results obtained in the previous experiment can be prevented just applying dropout at the inter-layer connections but not at the recurrent step.
+Taking into account that LSTMs solve the vanishing gradient problem using an internal state controlled by gates, it does not sound right to apply dropout over those gates.
+Probably the bad results obtained in the previous experiment can be prevented just applying dropout at the inter-layer connections but not at the recurrent step.
 
-In this experiment we apply different dropout values---25\%, 50\% and 75\%---after each LSTM layer.
+In this experiment we apply different dropout factors---25\%, 50\% and 75\%---after each LSTM layer.
 
@@
 	\end{tabular}
-	\caption{My caption}
-	\label{my-label}
+	\caption{Results after testing different dropout factors, only between layers.}
+	\label{t:drop3}
 \end{table}
@@
 	\includegraphics[width=\textwidth]{drop31}
-	\caption{TODO}
+	\caption{Validation and training loss with each dropout factor, for input sizes of 6 samples, only between layers.}
 	\label{f:drop31}
@@
 	\includegraphics[width=\textwidth]{drop32}
-	\caption{TODO}
+	\caption{Validation and training loss with each dropout factor, for input sizes of 48 samples, only between layers.}
 	\label{f:drop32}
@@
 
-As we can see, dropout does make any considerable improvement on the case in which the input size is 6---probably because that model was not overfitting the training data---; however, when the input size is 48, we can see how a dropout of 25\% successfully reduces the gap between training and validation losses, obtaining an overall lower test error.
+As we can see, dropout does make any considerable improvement when the input size is 6---probably because that model was not overfitting the training data---.
+However, when the input size is 48, we can see how a dropout of 25\% successfully reduces the gap between training and validation losses, obtaining an overall lower test error.
 
-In any case, the model with the bigger window size still has a worse performance than the model with an input size of 6. In the following experiment we only test this last model without dropout.
+In any case, this model still has a worse performance than the model with an input size of 6.
 
@@
 
-LSTM networks can learn the inherent data relations between over intervals of time of \emph{any} length; however, in our current architecture, we are feeding the model with short subsequences generated with a small sliding window.
+LSTM networks can learn the inherent data relations over intervals of time of \emph{any} length; however, in our current architecture, we train and test the model with short subsequences generated with a small sliding window.
 
 Why are we doing that?
-Can we train our LSTM with all our data set as a single sequence?
-If we do that, the LSTM network will learn whatever it finds relevant.
-It may happen that it finds data relations over large intervals of time, improving the overall performance of the system.
+Is it possible to train the LSTM model with all the data set expressed as a single huge sequence?
+By doing that, the LSTM network will learn whatever it finds relevant, and we will not arbitrarily decide its scope; the model will find the optimum one.
+It may happen that it finds data relations over very large intervals of time---maybe weeks---, improving the overall performance of the system.
 
-In our current architecture, if we try to process our data set with a huge window size, we find a physical problem.
+In our current architecture, if we try to process our data set with a huge window size, we find a physical problem: there is not enough RAM.
 
@@
 
-The memory used by this architecture directly depends on the size of sequences we use for training, making it not scalable for training with very large sequences.
+The memory used by this architecture is directly related to the size of sequences we use for training, making it not scalable for training with very large sequences of samples.
 We would need large amounts of RAM to exploit the true power of LSTM networks.
 
-To overcome this obstacle, in this experiment we present a new architecture with a model that gives a prediction for each input we feed to it (i.e.
+To overcome this obstacle, in this experiment we present a new architecture consisting in a sequence-to-sequence model that gives one prediction for each input we feed to it (i.e.
 when we feed a sequence of $ N $ inputs, the model gives $ N $ outputs).
@@
 
-In this new architecture each element of the training sequences pertain to different batches and we feed them one by one, hence the RAM used by the training process is constant and it does not depend on the sequences size.
-In fact, now the required RAM only depends on the batch size we choose.
+In this new architecture each element of the training sequences pertain to different batches, hence the RAM used by the training process does not depend on the sequences size anymore.
+Now the required RAM only depends on the batch size we choose.
 
-Our data is not compatible with this new training process and we must make a conversion first, since it is sequentially sorted by time.
-This conversion consists in reordering our data in such a way that, for a data set of $ N $ samples and a batch size of $ B $, samples from $ s_1 $ to $ s_{\frac{N}{B}} $ end up being the first elements of every batch, in sequential order; then samples from $ s_{\frac{N}{B}+1} $ to $ s_{\frac{N}{B}+\frac{N}{B}} $ end up being the second elements of every batch and so on.
-See figure \ref{f:explanation} for more details.
+Our current data set, since it is sequentially sorted by time, is not compatible with this new training process and we must make a conversion first.
+This conversion consists in reordering the measurements in such a way that, for a data set of $ N $ samples and a batch size of $ B $, samples from $ s_1 $ to $ s_{\frac{N}{B}} $ end up being the 1st element of every batch, in sequential order; then samples from $ s_{\frac{N}{B}+1} $ to $ s_{\frac{N}{B}+\frac{N}{B}} $ end up being the 2nd element of every batch and so on.
 
-TODO EXPLANATION
+As we can see, with this conversion we are actually generating $ B $ different training sequences.
+We can decide their size: with a batch size of 1 we train our model with a single sequence of 122,736 samples; with a batch size of 4 we train our model 4 sequences of size 30,684.
 
-As we can see, with this conversion we are training our model with $ B $ different sequences.
-Here we can decide how long we want our sequences: with a batch size of 1 we will train our model with a single sequence of 122,736 samples; with a batch size of 4 we will train our model 4 sequences of size 30,684.
+In this experiment we leave the batch size as 1000---so we train with input sequences of 122 samples---, in order to compare the results with the ones obtained in the previous experiments.
+We prefer to leave the same batch size because it directly affects to the model’s generalization, and results may vary.
 
-In this experiment we leave the batch size as 1000---we train with input sequences of 122 samples---, in order to compare the results with the ones obtained in the previous experiments.
-We must leave the same batch size because it directly affects to the model’s generalization.
-
-In this experiment we also clip all parameter gradients to a maximum norm of 0.000000001.
-Unfortunately we still suffer from exploding gradients, as it can be seen in figure \ref{f:architecture1}.
-This strict clipping slows the learning process; note that the gradients explode after after 2964 epochs---1 hour and 39 minutes---and the training loss is not even below 0.8.
+In this experiment we also clip all parameter gradients to a maximum norm of 0.000000001---very, very small---and even with that we still suffer from exploding gradients, as it can be seen in figure \ref{f:architecture1}.
+In addition, this strict clipping slows the learning process: note that the gradients explode after after 2964 epochs---1 hour and 39 minutes in our hardware---and the training loss is not even below 0.8.
 We could test lower clipping norms but then the training process would take an unfeasible amount of time to finish.
@@
 	\includegraphics[width=0.5\textwidth]{architecture1}
-	\caption{TODO}
+	\caption{Exploding gradients when trying to train the sequential model.}
 	\label{f:architecture1}
@@
 
-Our gradients are exploding due to the usage of ReLU, since its range is $ [0, +\inf] $.
-If we replace it with the hyperbolic tangent function---range $ [-1, 1] $---then we do not suffer from this problem.
+Our gradients are exploding due to the usage of ReLU as activation function, since its range is $ [0, +\infty) $.
+If we replace it with the hyperbolic tangent---range $ (-1, 1) $---then we do not suffer this problem.
 
@@
 	\end{tabular}
-	\caption{My caption}
-	\label{architecture2}
+	\caption{Results of the sliding window and the sequential model.}
+	\label{t:architecture2}
 \end{table}
@@
 	\includegraphics[width=0.5\textwidth]{architecture2}
-	\caption{TODO}
+	\caption{Validation and training loss of the sliding window and the sequential model.}
 	\label{f:architecture2}
@@
 This new architecture may be useful for problems where the model needs to recall information far from the past (e.g., stock price prediction or paragraph understanding) but it does not offer any advantage over our previous architecture based on a sliding window.
-From the previous experiments, we already had the suspicion that all the relevant information for making a good prediction relies on the short past.
+From the previous experiments, we already had the suspicion that all the relevant information for making a good prediction relies on the very short past.
 
@@
 
+In this practice we confirmed several expectations but also found cases where established methodologies have no effect in our context.
+
+For example, regarding to the network topology, since two hidden layers can approximate any smooth variable relation to any accuracy, we found that our model experiments an improvement when we go from one hidden layer to two, and we do not experiment any improvement when using three.
+We could apply some deep learning techniques, such as layer-wise pre-training deep autoencoders with several layers, but it was out of the scope of this RNN practice.
+
+Regarding to the activation function, as found by Clevert et al, 2016 and confirmed by us in the prior DL-MAI practice, we expected to speed up the training process and obtain a lesser error with ELUs.
+However, in this case, the change had no effect.
+
+We have the suspicion that the cause behind this behavior relies on the data we are working with: easy to approximate, no matter the activation function, but with a constant underlaying noise, impossible to predict with the variables we have.
+With more variables, this noise can be reduced.
+
+The rest of the activation functions we tested (logistic, hyperbolic tangent, softplus) obtain similar results than ReLU, confirming our suspicion.
+
+Now, regarding to specific RNN properties, we found that larger window sizes do not improve the performance of the system, even when regularizing overfited models using dropout.
+Input sequences of 6 steps are enough to predict with the maximum accuracy we found the following 6.
+In fact, we also found that our alternative architecture consisting in a sequence-to-sequence model trained with very long sequences is not able to beat that mark neither.
+
+Finally, regarding to regularization, we tested dropout and we found that in our case it is better not to apply it on the recursive step of the network, since the network performance is dramatically reduced.
+We suspect that dropout may interfere with the underlying state of the LSTM units, causing them to forget important information.
+If the model overfits, in our case, it is better to apply dropout between layers.
+
 \section{Conclusion and Future Work}
 
+In this practice we extended the experiments performed in the previous ``Guided Laboratory of the RNN Topic'', predicting more steps, trying more complex topologies and testing a sequence-to-sequence model.
+
+We presented and followed a plan of experiments that lead us to confirm and discard several hypothesis, as explained in the prior section.
+
+Due to the nature of the data set and the results of the performed experiments, in this practice we did not test different optimizers or regularization methods.
+Due to lack of time, we have not tried our model against other sites either, and we did not check the variance of our error.
+It is recommendable to perform n-fold cross-validation instead of a single test.
+
+Finally, it would be also interesting to find a data set where our sequence-to-sequence architecture may obtain better results than the sliding window process.
+As previously said, this model can be useful on cases where long-term relations have special importance, such as on stock pricing prediction or text understanding.
+
+\bibliography{bibfile}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2017-12-07 process.py

@@ -59,7 +59,7 @@ def plotOptions(results, title, ylabel, keys):
     plt.ylabel(ylabel)
     plt.xlabel('Epoch')
     
-    plt.ylim(ymin=0.02, ymax=0.06)
+    plt.ylim(ymin=0.02, ymax=0.08)
     
     plt.show()
     

### 2017-12-08 baseline.py
All experiments performed

@@ -1,0 +1,243 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation
+from keras.layers import LSTM, GRU, Dropout
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+
+__author__ = 'bejar, lao'
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    plt.ylim(ymin=0.025, ymax=0.037)
+    
+    plt.show()
+
+def lagged_vector(data, lag=1):
+    """
+    Returns a matrix with columns that are the steps of the lagged time series
+    Last column is the value to predict
+    :param data:
+    :param lag:
+    :return:
+    """
+    lvect = []
+    for i in xrange(lag):
+        lvect.append(data[i: -lag+i])
+    lvect.append(data[lag:])
+    return np.stack(lvect, axis=1)
+
+def load_config_file(nfile, abspath=False):
+    """
+    Read the configuration from a json file
+
+    :param abspath:
+    :param nfile:
+    :return:
+    """
+    ext = '.json' if 'json' not in nfile else ''
+    pre = '' if abspath else './'
+    fp = open(pre + nfile + ext, 'r')
+
+    s = ''
+
+    for l in fp:
+        s += l
+
+    return json.loads(s)
+
+if __name__ == '__main__':
+    start = time.time()
+    
[…]

### 2017-12-08 multistep.py

@@ -1,0 +1,297 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation
+from keras.layers import LSTM, GRU, Dropout
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+from keras.models import load_model
+
+__author__ = 'bejar, lao'
+
+DELAYS = 55
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    #plt.ylim(ymin=0.025, ymax=0.032)
+    
+    plt.show()
+
+def plotMSE(results, title):
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['mseEvolution'][:DELAYS])
+    
+    plt.legend(results.keys(), loc='upper left')
+    
+    plt.title(title)
+    plt.ylabel('MSE')
+    plt.xlabel('Unknown steps')
+    
+    plt.show()
+
+def lagged_vector(data, lag=1):
+    """
+    Returns a matrix with columns that are the steps of the lagged time series
+    Last column is the value to predict
+    :param data:
+    :param lag:
+    :return:
+    """
+    lvect = []
+    for i in xrange(lag):
+        lvect.append(data[i: -lag+i])
+    lvect.append(data[lag:])
+    return np.stack(lvect, axis=1)
+
[…]

### 2017-12-08 process.py

@@ -1,291 +1,0 @@
-"""
-.. module:: WindPrediction
-
-WindPrediction
-*************
-
-:Description: WindPrediction
-
-:Authors: bejar, lao
-    
-
-:Version: 
-
-:Created on: 06/09/2017 9:47
-
-:Updated on: 22/11/2017 0:20 
-
-"""
-
-import numpy as np
-
-from keras.models import Sequential
-from keras.layers import Dense, Activation, Dropout
-from keras.layers import LSTM, GRU
-from keras.optimizers import RMSprop, SGD
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import StandardScaler
-from sklearn.metrics import mean_squared_error
-import argparse
-import json
-import time
-from keras.callbacks import EarlyStopping
-import matplotlib.pyplot as plt
-import os
-import pickle
-import sys
-from keras.utils import plot_model
-
-__author__ = 'bejar, lao'
-
-FUTURE = 6
-
-def plotOptions(results, title, ylabel, keys):
-    plt.gca().set_color_cycle(None)
-    plt.plot([], '--', color='black')
-    plt.plot([], color='black')
-    
-    for i in results:
-        plt.plot(results[i]['h'][keys[0]])
-    
-    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
-    
-    plt.gca().set_color_cycle(None)
-    
-    for i in results:
-        plt.plot(results[i]['h'][keys[1]], '--')
-    
-    plt.title(title)
-    plt.ylabel(ylabel)
-    plt.xlabel('Epoch')
-    
-    plt.ylim(ymin=0.02, ymax=0.08)
-    
-    plt.show()
-    
-# Makes the dataset compatible with keras stateful LSTM's, where the ith element of the batch uses the state stored for the ith element of the previous batch.
-def batchify((x, y), batchSize):
-    x, y = removeSurplus((x, y), batchSize)
-    x = batchifyImpl(x, batchSize)
-    y = batchifyImpl(y, batchSize)
-    return x, y
-
-def batchifyImpl(data, batchSize):
-    n = len(data)
-    parsedData = np.empty(data.shape)
-    k = 0
-    
-    for i in xrange(batchSize):
-        for j in xrange(i, n, batchSize):
-            parsedData[k] = data[j]
-            k += 1
-    
-    return parsedData
-
-def removeSurplus((x, y), batchSize):
-    surplus = x.shape[0] % batchSize
-    
-    if surplus == 0:
[…]

### 2017-12-08 seq2seq.py

@@ -1,0 +1,258 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation, RepeatVector, TimeDistributed
+from keras.layers import LSTM, GRU, Dropout
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+from scipy.optimize._differentialevolution import differential_evolution
+
+__author__ = 'bejar, lao'
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    plt.show()
+
+def plotMSE(results, title):
+    plt.gca().set_color_cycle(None)
+    
+    evolution = sorted([[results[i]['future'], results[i]['test_loss']] for i in results], key=lambda d: d[0])
+    x = [d[0] for d in evolution]
+    y = [d[1] for d in evolution]
+    
+    plt.plot(x, y)
+    
+    plt.title(title)
+    plt.ylabel('MSE')
+    plt.xlabel('Predictions')
+    
+    plt.show()
+
+def lagged_vector(data, lag=1):
+    """
+    Returns a matrix with columns that are the steps of the lagged time series
+    Last column is the value to predict
+    :param data:
+    :param lag:
+    :return:
+    """
+    lvect = []
+    for i in xrange(lag):
+        lvect.append(data[i: -lag+i])
+    lvect.append(data[lag:])
[…]

### 2017-12-08 sites.py

@@ -1,0 +1,243 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation
+from keras.layers import LSTM, GRU, Dropout
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+
+__author__ = 'bejar, lao'
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    plt.ylim(ymin=0.025, ymax=0.037)
+    
+    plt.show()
+
+def lagged_vector(data, lag=1):
+    """
+    Returns a matrix with columns that are the steps of the lagged time series
+    Last column is the value to predict
+    :param data:
+    :param lag:
+    :return:
+    """
+    lvect = []
+    for i in xrange(lag):
+        lvect.append(data[i: -lag+i])
+    lvect.append(data[lag:])
+    return np.stack(lvect, axis=1)
+
+def load_config_file(nfile, abspath=False):
+    """
+    Read the configuration from a json file
+
+    :param abspath:
+    :param nfile:
+    :return:
+    """
+    ext = '.json' if 'json' not in nfile else ''
+    pre = '' if abspath else './'
+    fp = open(pre + nfile + ext, 'r')
+
+    s = ''
+
+    for l in fp:
+        s += l
+
+    return json.loads(s)
+
+if __name__ == '__main__':
[…]

### 2017-12-08 variables.py

@@ -1,0 +1,243 @@
+"""
+.. module:: WindPrediction
+
+WindPrediction
+*************
+
+:Description: WindPrediction
+
+:Authors: bejar, lao
+    
+
+:Version: 
+
+:Created on: 06/09/2017 9:47 
+
+"""
+
+import numpy as np
+
+from keras.models import Sequential
+from keras.layers import Dense, Activation
+from keras.layers import LSTM, GRU, Dropout
+from keras.optimizers import RMSprop, SGD
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import StandardScaler
+from sklearn.metrics import mean_squared_error
+import argparse
+import json
+import time
+from keras.callbacks import EarlyStopping
+import matplotlib.pyplot as plt
+import os
+import pickle
+import sys
+from keras.utils import plot_model
+
+__author__ = 'bejar, lao'
+
+def plotOptions(results, title, ylabel, keys):
+    plt.gca().set_color_cycle(None)
+    plt.plot([], '--', color='black')
+    plt.plot([], color='black')
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[0]])
+    
+    plt.legend(['Training', 'Validation'] + results.keys(), loc='upper right')
+    
+    plt.gca().set_color_cycle(None)
+    
+    for i in results:
+        plt.plot(results[i]['h'][keys[1]], '--')
+    
+    plt.title(title)
+    plt.ylabel(ylabel)
+    plt.xlabel('Epoch')
+    
+    plt.ylim(ymin=0.024, ymax=0.045)
+    
+    plt.show()
+
+def lagged_vector(data, lag=1):
+    """
+    Returns a matrix with columns that are the steps of the lagged time series
+    Last column is the value to predict
+    :param data:
+    :param lag:
+    :return:
+    """
+    lvect = []
+    for i in xrange(lag):
+        lvect.append(data[i: -lag+i])
+    lvect.append(data[lag:])
+    return np.stack(lvect, axis=1)
+
+def load_config_file(nfile, abspath=False):
+    """
+    Read the configuration from a json file
+
+    :param abspath:
+    :param nfile:
+    :return:
+    """
+    ext = '.json' if 'json' not in nfile else ''
+    pre = '' if abspath else './'
+    fp = open(pre + nfile + ext, 'r')
+
+    s = ''
+
+    for l in fp:
+        s += l
+
+    return json.loads(s)
+
+if __name__ == '__main__':
+    start = time.time()
+    
[…]

### 2017-12-08 paper/paper.tex
Paper finished

\documentclass[]{article}

\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{float}
\usepackage{hyperref}

\title{Recurrent Neural Networks}
\author{Juan Lao Tebar}

\begin{document}

	\maketitle

	\section{Introduction}

	In this practice we explore the effects of the different elements of RNNs using a specific data set\footnote{https://upc-mai-dl.github.io/rnn-lab-autonomous/}. In this case we extend the Wind prediction example that we used in the previous guided session.

	In the original example we predicted one step of the future wind---next 15 minutes---using a window of previous measurements. Here we extend the problem in different ways:

	\begin{description}
		\item [Complementary variables] In the original experiment we used only the wind speed as input data. Now we include also the air density, the temperature and the pressure.

		\item [Complementary sites] In the original experiment we predicted the wind speed using the data from one site, but the data set includes three additional sites that are geographically close (they are in the vertices of a 2 km square). Here we leverage the information of the wind speed from the other sites.

		\item [Multi step prediction] Here we obtain a multi step prediction using the original model simply by adding the value predicted to the current set of measurements and discarding the oldest one.

		\item [Sequence to sequence prediction] An alternative to shifting the input and adding the predictions to have a multi step prediction is a network that links a window of measurements to a window of predictions.
	\end{description}

	In this practice we change the stopping criteria of the training process, as we remove the limit of epochs and we implement a stopper based on the validation loss: when we detect a minimum value, we stop the training after 30 epochs if it does not decrease. We consider that we should not stop the training process just because we reached an arbitrary number of epochs if the model does not show signs of convergence.

	In the guided laboratory we trained the model with 50,000 samples, validated with 10,000 and tested with 10,000. In this practice we use the whole data set, training our model with 122,736 samples, validating with 61,368 and testing with 61,368 for no apparent reason\footnote{Actually, the reason behind this decision is just that we performed some dummy tests using the whole data set and we forgot to set the default value again.}.

	The code related to this work is public and available at github\footnote{https://github.com/juanlao7/DL-lab2}.

	\section{Baseline}

	Before performing any experiment, we must define a baseline model in order to compare the results and determine how every change affects the accuracy of the system.

	Our baseline model is similar to the one proposed in the guided laboratory: input window size of 6 steps, 1 output, 512 LSTM units in one single hidden layer and no dropout.

	Table \ref{t:baseline} shows the results obtained with this model, while figure \ref{f:baseline} shows the evolution of the training and validation loss.

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}ccc@{}}
			\toprule
			Training loss & Validation loss & Test loss \\ \midrule
			0.02729       & 0.02924         & 0.03339   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained with the baseline model.}
		\label{t:baseline}
	\end{table}

	\section{Complementary Variables}

	\subsection{Plan}

	In this session we modify the architecture of the system adequately for taking all the variables as input---but still giving only one variable as output---and we study how the MSE changes when we change the size of hidden layer, when we apply dropout and when we change the size of input window.

	\begin{itemize}
		\item In the 1st experiment we test different layer sizes: 32, 128, 512 and 1024 units.

		\item In the 2nd experiment we test different dropout factors applied in the LSTM layer, including its gates: 0\%, 25\%, 50\%, 75\%.

		\item In the 3rd experiment we test different dropout factors only applied on inter-layer connections, between the hidden and the output layer: 0\%, 25\%, 50\%, 75\%.

		\item In the 4th experiment we test different input window sizes: 1, 3, 6, 12, 24, 48. In this experiment, in the case of a window size of 48, we had to clip all parameter gradients to a maximum norm of 1 in order to prevent exploding gradients.
	\end{itemize}

	\subsection{Expectations}

	Regarding to learning techniques:

	\begin{itemize}
		\item In theory, adding more units to the hidden layer leads to better training accuracies. However, the model may lose power of generalization, overfitting the training data and offering worse validation and test accuracies.

		\item If the network overfits the training data, we expect that applying a dropout factor will successfully regularize the model. However, the dropout implementation we use---``dropout'' argument in Keras implementation of LSTM layers---affects also to the LSTM gates, thus it is possible that the model will end up performing even worse due to the vanishing gradient problem. For this reason we also test an inter-layer dropout, that does not affect to LSTM gates. On the other hand, if the model does not overfit, dropout will slightly reduce the gap between training and validation losses, but not in a considerable way.

		\item We suspect that in this data set there are no data relations over long intervals of time, hence we will not obtain a better result by increasing the window size. However, decreasing it may lead to a worse accuracy.
	\end{itemize}

	Regarding to this specific problem:

	\begin{itemize}
		\item More information---variables---are always helpful and the model may find new inter-variable relations that lead to a better accuracy.
	\end{itemize}

	\subsection{Results}

	\begin{itemize}
		\item Table \ref{t:variables_1} shows the results obtained after testing different layer sizes, while figure \ref{f:variables_1} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:variables_2} shows the results obtained after testing different dropout factors, while figure \ref{f:variables_2} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:variables_3} shows the results obtained after testing different inter-layer dropout factors, while figure \ref{f:variables_3} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:variables_4} shows the results obtained after testing different input window sizes, while figures \ref{f:variables_4_1} and \ref{f:variables_4_2} show the evolution of the training and validation loss in each case.
	\end{itemize}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			32 LSTM units   & 0.02770       & 0.02939         & 0.03327   \\
			128 LSTM units  & 0.02621       & 0.02835         & 0.03218   \\
			512 LSTM units  & 0.02562       & 0.02911         & 0.03266   \\
			1024 LSTM units & 0.02531       & 0.02883         & 0.03247   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different layer sizes.}
		\label{t:variables_1}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{variables_1}
		\caption{Validation and training loss with each layer size.}
		\label{f:variables_1}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			0\% dropout  & 0.02562       & 0.02911         & 0.03266   \\
			25\% dropout & 0.24172       & 0.08519         & 0.08502   \\
			50\% dropout & 0.47773       & 0.27423         & 0.26216   \\
			75\% dropout & 0.73191       & 0.57920         & 0.54833   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different dropout factors.}
		\label{t:variables_2}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{variables_2}
		\caption{Validation and training loss with each dropout factor.}
		\label{f:variables_2}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			0\% inter-layer dropout  & 0.02562       & 0.02911         & 0.03266   \\
			25\% inter-layer dropout & 0.02830       & 0.02886         & 0.03258   \\
			50\% inter-layer dropout & 0.03110       & 0.02884         & 0.03262   \\
			75\% inter-layer dropout & 0.04012       & 0.02975         & 0.03337   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different inter-layer dropout factors.}
		\label{t:variables_3}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{variables_3}
		\caption{Validation and training loss with each inter-layer dropout factor.}
		\label{f:variables_3}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			Window size of 1  & 0.03146       & 0.03305         & 0.03599   \\
			Window size of 3  & 0.02690       & 0.02880         & 0.03254   \\
			Window size of 6  & 0.02562       & 0.02911         & 0.03266   \\
			Window size of 12 & 0.02586       & 0.02884         & 0.03260   \\
			Window size of 24 & 0.02465       & 0.02941         & 0.03375   \\
			Window size of 48 & 0.02543       & 0.02885         & 0.03316   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different input window sizes.}
		\label{t:variables_4}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{variables_4_1}
		\caption{Validation and training loss with window sizes 1, 3 and 6.}
		\label{f:variables_4_1}
	\end{figure}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{variables_4_2}
		\caption{Validation and training loss with window sizes 6, 12, 24 and 48.}
		\label{f:variables_4_2}
	\end{figure}

	\subsection{Conclusions}

	Regarding to learning techniques:

	\begin{itemize}
		\item \emph{As expected}, adding more units to the hidden layer leads to better training accuracies and, for the case of 1024 units, overfits the training data and offers worse validation and test accuracies.

		\item \emph{As expected}, dropout factors applied to LSTM gates leads to far worse results than the original approach. In our case, it is not a good idea to mess with the current state of the units.

		\item \emph{As expected}, since our model does not overfit the training data, certain dropout factors---25\% in this case---reduce the gap between the training and validation loss, but does not increase the accuracy of the system.

		\item \emph{As expected}, smaller window sizes lead to worse accuracies while bigger window sizes do not offer much improvement. However, looks like the model converges in a less number of epochs.
	\end{itemize}

	Regarding to this specific problem:

	\begin{itemize}
		\item \emph{As expected}, more information---variables---are helpful and the model finds new inter-variable relations that lead to a \emph{slightly} better accuracy. The baseline model got a test accuracy of 0.03339 while our best model here got 0.03218---and several configurations got a mark under 0.03290---. We can conclude that air density, pressure and temperature help to predict wind speed at 100m in site ``90-45142''.
	\end{itemize}

	\section{Complementary Sites}

	\subsection{Plan}

	In this session we leverage the information of the wind speed from the other sites and we study how the MSE changes when we change the size of hidden layer, when we apply dropout and when we change the size of input window.

	\begin{itemize}
		\item In the 1st experiment we test different layer sizes: 32, 128, 512 and 1024 units.

		\item In the 2nd experiment we test different dropout factors applied in the LSTM layer, including its gates: 0\%, 25\%, 50\%, 75\%.

		\item In the 3rd experiment we test different dropout factors only applied on inter-layer connections, between the hidden and the output layer: 0\%, 25\%, 50\%, 75\%.

		\item In the 4th experiment we test different input window sizes: 1, 3, 6, 12, 24, 48. In this experiment, in the case of a window size of 48, we had to clip all parameter gradients to a maximum norm of 1 in order to prevent exploding gradients.
	\end{itemize}

	\subsection{Expectations}

	Exactly the same as in section Complementary Variables.

	\subsection{Results}

	\begin{itemize}
		\item Table \ref{t:sites_1} shows the results obtained after testing different layer sizes, while figure \ref{f:sites_1} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:sites_2} shows the results obtained after testing different dropout factors, while figure \ref{f:sites_2} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:sites_3} shows the results obtained after testing different inter-layer dropout factors, while figure \ref{f:sites_3} shows the evolution of the training and validation loss in each case.

		\item Table \ref{t:sites_4} shows the results obtained after testing different input window sizes, while figures \ref{f:sites_4_1} and \ref{f:sites_4_2} show the evolution of the training and validation loss in each case.
	\end{itemize}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			32 LSTM units   & 0.02666       & 0.02821         & 0.03231   \\
			128 LSTM units  & 0.02694       & 0.02855         & 0.03246   \\
			512 LSTM units  & 0.02631       & 0.02847         & 0.03225   \\
			1024 LSTM units & 0.02591       & 0.02882         & 0.03242   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different layer sizes.}
		\label{t:sites_1}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{sites_1}
		\caption{Validation and training loss with each layer size.}
		\label{f:sites_1}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			0\% dropout  & 0.02631       & 0.02847         & 0.03225   \\
			25\% dropout & 0.03554       & 0.08211         & 0.08097   \\
			50\% dropout & 0.09668       & 0.26654         & 0.25337   \\
			75\% dropout & 0.34805       & 0.54228         & 0.51081   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different dropout factors.}
		\label{t:sites_2}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{sites_2}
		\caption{Validation and training loss with each dropout factor.}
		\label{f:sites_2}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			0\% inter-layer dropout  & 0.02631       & 0.02847         & 0.03225   \\
			25\% inter-layer dropout & 0.02836       & 0.02887         & 0.03240   \\
			50\% inter-layer dropout & 0.03084       & 0.02876         & 0.03277   \\
			75\% inter-layer dropout & 0.03898       & 0.02872         & 0.03251   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different inter-layer dropout factors.}
		\label{t:sites_3}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{sites_3}
		\caption{Validation and training loss with each inter-layer dropout factor.}
		\label{f:sites_3}
	\end{figure}

	\begin{table}[H]
		\centering
		\begin{tabular}{@{}cccc@{}}
			\toprule
			& Training loss & Validation loss & Test loss \\ \midrule
			Window size of 1  & 0.03136       & 0.03277         & 0.03559   \\
			Window size of 3  & 0.02669       & 0.02843         & 0.03226   \\
			Window size of 6  & 0.02631       & 0.02847         & 0.03225   \\
			Window size of 12 & 0.02558       & 0.02935         & 0.03302   \\
			Window size of 24 & 0.02571       & 0.02879         & 0.03264   \\ \bottomrule
		\end{tabular}
		\caption{Results obtained after testing different input window sizes.}
		\label{t:sites_4}
	\end{table}

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{sites_4_1}
		\caption{Validation and training loss with window sizes 1, 3 and 6.}
		\label{f:sites_4_1}
	\end{figure}

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{sites_4_2}
		\caption{Validation and training loss with window sizes 6, 12, 24 and 48.}
		\label{f:sites_4_2}
	\end{figure}

	\subsection{Conclusions}

	Regarding to learning techniques: exactly the same as in section Complementary Variables.

	Regarding to this specific problem:

	\begin{itemize}
		\item \emph{As expected}, more information---variables---are helpful and the model finds new inter-variable relations that lead to a \emph{slightly} better accuracy. The baseline model got a test accuracy of 0.03339 while our best model here got 0.03225---and several configurations got a mark under 0.03290---. We can conclude that wind speeds at 100m of sites ``90-45143', ``90-45229'' and ``90-45230'' help to predict wind speed at 100m in site ``90-45142''.
	\end{itemize}

	\section{Multi Step Prediction}

	\subsection{Plan}

	In this session we analyze the behavior of the model when it is fed with its own outputs as new inputs; we study how the MSE degrades the more steps in the future we predict and how the MSE changes if we extend the input window.

	In this experiment we test different input window sizes---1, 3, 6, 12, 24---and we observe the evolution of MSE when trying to predict the next 55 steps.

	\subsection{Expectations}

	\begin{itemize}
		\item In previous experiments we found that small window sizes lead to a higher error, and big window sizes do not show any improvement. We expect that the error of the model accumulates as more steps in the future we predict, leading to a higher error on small window sizes and more stable accuracies in big window sizes.
	\end{itemize}

	\subsection{Results}

	Figure \ref{f:multistep} shows the results obtained after testing different input window sizes.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.5\textwidth]{multistep}
		\caption{MSE evolution for each window size.}
		\label{f:multistep}
	\end{figure}

	\subsection{Conclusions}

	\begin{itemize}
		\item \emph{Surprisingly}, we found the opposite of our expectations. As we can see in figure \ref{f:multistep}, with small window sizes the error seems to converge while in big window sizes the error grows exponentially. Sincerely, we do not have an explanation for this interesting phenomena.
	\end{itemize}

	\section{Sequence to Sequence Prediction}

	\subsection{Plan}

	In this session we modify the architecture of the system adequately for linking a window of measurements to a window of predictions, and we study how the MSE of the predictions changes with the length of the predicted sequence.

	In this experiment we test different prediction lengths---1, 3, 6, 12, 24---and we observe the evolution of the MSE.

	\subsection{Expectations}

	\begin{itemize}
		\item Models that try to predict more steps forward will lead to lower accuracies.
		\item This sequence-to-sequence model will have a lower error than the previous multi step approach, since this model does not accumulates the error.
	\end{itemize}

	\subsection{Results}

	Figure \ref{f:seq2seq_1} shows the evolution of the training and validation loss in each case, while figure \ref{f:seq2seq_2} shows the MSE evolution.

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{seq2seq_1}
		\caption{Validation and training loss with each number of predictions.}
		\label{f:seq2seq_1}
	\end{figure}

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{seq2seq_2}
		\caption{MSE evolution relation with the number of predictions.}
		\label{f:seq2seq_2}
	\end{figure}

	\subsection{Conclusions}

	\begin{itemize}
		\item \emph{As expected}, models that try to predict more steps forward lead to lower accuracies.
		\item \emph{As expected}, this sequence-to-sequence model has a lower error than the previous multi step approach. The previous model, with a window size of 6 and 24 unknown inputs, has a MSE of 1.06895 while this model has a MSE of 0.384457 when predicting 24 steps.
	\end{itemize}

\end{document}

### 2017-12-08 multistep.py

@@ -257,6 +257,8 @@ if __name__ == '__main__':
             handler = open(mseFileName, 'rb')
             results[configName]['mseEvolution'] = pickle.load(handler)
             handler.close()
+            
+            print results[configName]['mseEvolution'][24]
         else:
             mseEvolution = []
             

### 2017-12-10 seq2seq.py
fixes imports

@@ -33,7 +33,6 @@ import os
 import pickle
 import sys
 from keras.utils import plot_model
-from scipy.optimize._differentialevolution import differential_evolution
 
 __author__ = 'bejar, lao'
 

### 2017-12-10 paper/paper.tex
"data set" -> "dataset"

@@
 
-	In this practice we explore the effects of the different elements of RNNs using a specific data set\footnote{https://upc-mai-dl.github.io/rnn-lab-autonomous/}.
+	In this practice we explore the effects of the different elements of RNNs using a specific dataset\footnote{https://upc-mai-dl.github.io/rnn-lab-autonomous/}.
 In this case we extend the Wind prediction example that we used in the previous guided session.
@@
 
-		\item [Complementary sites] In the original experiment we predicted the wind speed using the data from one site, but the data set includes three additional sites that are geographically close (they are in the vertices of a 2 km square).
+		\item [Complementary sites] In the original experiment we predicted the wind speed using the data from one site, but the dataset includes three additional sites that are geographically close (they are in the vertices of a 2 km square).
 Here we leverage the information of the wind speed from the other sites.
@@
 	In the guided laboratory we trained the model with 50,000 samples, validated with 10,000 and tested with 10,000.
-In this practice we use the whole data set, training our model with 122,736 samples, validating with 61,368 and testing with 61,368 for no apparent reason\footnote{Actually, the reason behind this decision is just that we performed some dummy tests using the whole data set and we forgot to set the default value again.}.
+In this practice we use the whole dataset, training our model with 122,736 samples, validating with 61,368 and testing with 61,368 for no apparent reason\footnote{Actually, the reason behind this decision is just that we performed some dummy tests using the whole dataset and we forgot to set the default value again.}.
 
@@
 
-		\item We suspect that in this data set there are no data relations over long intervals of time, hence we will not obtain a better result by increasing the window size.
+		\item We suspect that in this dataset there are no data relations over long intervals of time, hence we will not obtain a better result by increasing the window size.
 However, decreasing it may lead to a worse accuracy.

### 2017-12-10 paper/paper.tex
small changes

@@
 
-	Our baseline model is similar to the one proposed in the guided laboratory: input window size of 6 steps, 1 output, 512 LSTM units in one single hidden layer and no dropout.
+	Our baseline model is similar to the one proposed in the guided laboratory: input window size of 6 steps, 1 output, 512 LSTM units in one single hidden layer and no dropout, trained with RMSprop---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay.
 
-	Table \ref{t:baseline} shows the results obtained with this model, while figure \ref{f:baseline} shows the evolution of the training and validation loss.
+	Table \ref{t:baseline} shows the results obtained with this model.
 

### 2017-12-11 paper/paper.tex
Fixes seq2seq model

@@
 
-	In this practice we explore the effects of the different elements of RNNs using a specific dataset\footnote{https://upc-mai-dl.github.io/rnn-lab-autonomous/}.
+	In this practice we explore the effects of the different elements of RNNs using a specific dataset\footnote{\url{https://upc-mai-dl.github.io/rnn-lab-autonomous/}}.
 In this case we extend the Wind prediction example that we used in the previous guided session.
@@
 
-	The code related to this work is public and available at github\footnote{https://github.com/juanlao7/DL-lab2}.
+	The code related to this work is public and available at github\footnote{\url{https://github.com/juanlao7/DL-lab2}}.
 
@@
 		\item \emph{As expected}, this sequence-to-sequence model has a lower error than the previous multi step approach.
-The previous model, with a window size of 6 and 24 unknown inputs, has a MSE of 1.06895 while this model has a MSE of 0.384457 when predicting 24 steps.
+The previous model, with a window size of 6 and 24 unknown inputs, has a MSE of 1.06895 while this model has a MSE of 0.368767 when predicting 24 steps.
 	\end{itemize}

### 2017-12-11 seq2seq.py

@@ -184,26 +184,17 @@ if __name__ == '__main__':
         activation_r = config['activation_r']
     
         model = Sequential()
-        if nlayers == 1:
-            model.add(RNN(neurons, input_shape=(train_x.shape[1], train_x.shape[2]), implementation=impl, dropout=drop,
-                          activation=activation, recurrent_activation=activation_r))
-        else:
-            model.add(RNN(neurons, input_shape=(train_x.shape[1], train_x.shape[2]), implementation=impl, dropout=drop,
-                          activation=activation, recurrent_activation=activation_r, return_sequences=True))
-            model.add(Dropout(interdrop))
-            
-            for i in range(1, nlayers-1):
-                model.add(RNN(neurons, dropout=drop, implementation=impl,
-                              activation=activation, recurrent_activation=activation_r, return_sequences=True))
-                model.add(Dropout(interdrop))
-                
-            model.add(RNN(neurons, dropout=drop, implementation=impl,
-                          activation=activation, recurrent_activation=activation_r))
         
-        
-        model.add(Dropout(interdrop))
+        model.add(RNN(neurons, input_shape=(train_x.shape[1], train_x.shape[2]), implementation=impl, dropout=drop,
+                          activation=activation, recurrent_activation=activation_r))
         
         model.add(RepeatVector(future))
+
+        for i in range(0, nlayers):
+            model.add(RNN(neurons, dropout=drop, implementation=impl,
+                          activation=activation, recurrent_activation=activation_r, return_sequences=True))
+            model.add(Dropout(interdrop))
+        
         model.add(TimeDistributed(Dense(1)))
     
         ############################################

### 2017-12-17 paper/paper.tex
Grammar corrections

@@
 
-	In this practice we explore the effects of the different elements of RNNs using a specific dataset\footnote{\url{https://upc-mai-dl.github.io/rnn-lab-autonomous/}}.
-In this case we extend the Wind prediction example that we used in the previous guided session.
+	In this practice we explore the effects of the different elements of RNNs using a specific dataset\footnote{\url{https://upc-mai-dl.github.io/rnn-lab-autonomous/}} by extending the Wind prediction example of the previous guided session.
 
@@
 	\begin{description}
-		\item [Complementary variables] In the original experiment we used only the wind speed as input data. Now we include also the air density, the temperature and the pressure.
+		\item [Complementary variables.] In the original experiment we used only the wind speed as input data. Now we include also the air density, the temperature and the pressure.
 
-		\item [Complementary sites] In the original experiment we predicted the wind speed using the data from one site, but the dataset includes three additional sites that are geographically close (they are in the vertices of a 2 km square).
+		\item [Complementary sites.] In the original experiment we predicted the wind speed using the data from one site, but the dataset includes three additional sites that are geographically close---they are in the vertices of a 2 km square---.
 Here we leverage the information of the wind speed from the other sites.
 
-		\item [Multi step prediction] Here we obtain a multi step prediction using the original model simply by adding the value predicted to the current set of measurements and discarding the oldest one.
+		\item [Multi step prediction.] Here we obtain a multi step prediction using the original model simply by adding the value predicted to the current set of measurements and discarding the oldest one.
 
-		\item [Sequence to sequence prediction] An alternative to shifting the input and adding the predictions to have a multi step prediction is a network that links a window of measurements to a window of predictions.
+		\item [Sequence to sequence prediction.] An alternative to shifting the input and adding the predictions to have a multi step prediction is a network that links a window of measurements to a window of predictions.
 	\end{description}
 
-	In this practice we change the stopping criteria of the training process, as we remove the limit of epochs and we implement a stopper based on the validation loss: when we detect a minimum value, we stop the training after 30 epochs if it does not decrease.
-We consider that we should not stop the training process just because we reached an arbitrary number of epochs if the model does not show signs of convergence.
+	In this practice we change the stopping criteria of the training process, as we remove the limit of epochs and we implement a stopper based on the validation loss: when we detect a minimum value, we stop the training after 30 epochs if it does not decrease again.
+If the model does not show any sign of convergence, we consider that we should not stop the training process just because it reached an arbitrary number of epochs.
 
-	In the guided laboratory we trained the model with 50,000 samples, validated with 10,000 and tested with 10,000.
-In this practice we use the whole dataset, training our model with 122,736 samples, validating with 61,368 and testing with 61,368 for no apparent reason\footnote{Actually, the reason behind this decision is just that we performed some dummy tests using the whole dataset and we forgot to set the default value again.}.
+	In the guided laboratory we trained the model with 50,000 samples, validated with 10,000 and tested with 10,000, while in this practice we use the whole dataset, training our model with 122,736 samples, validating with 61,368 and testing with 61,368.
 
@@
 
-	Before performing any experiment, we must define a baseline model in order to compare the results and determine how every change affects the accuracy of the system.
+	Before performing any experiment, we define a baseline model in order to compare the results and determine how each change affects the accuracy of the system.
 
-	Our baseline model is similar to the one proposed in the guided laboratory: input window size of 6 steps, 1 output, 512 LSTM units in one single hidden layer and no dropout, trained with RMSprop---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay.
+	Our baseline model is similar to the one proposed in the guided laboratory: an input window size of 6 steps, 1 output, 512 LSTM units in one single hidden layer and no dropout, trained with RMSprop---with a learning rate of 0.0001, $ \rho = 0.9 $, $ \varepsilon = 1 \cdot 10^{-8} $ and no learning rate decay.
 
@@
 
-	In this session we modify the architecture of the system adequately for taking all the variables as input---but still giving only one variable as output---and we study how the MSE changes when we change the size of hidden layer, when we apply dropout and when we change the size of input window.
+	In this session we modify the architecture of the system adequately for taking all the variables as input---but still giving only one variable as output---.
+Then we study how the MSE is affected by the size of the hidden layer, the dropout and the size of the input window.
 
@@
 		\item In the 4th experiment we test different input window sizes: 1, 3, 6, 12, 24, 48.
-In this experiment, in the case of a window size of 48, we had to clip all parameter gradients to a maximum norm of 1 in order to prevent exploding gradients.
+In this experiment, in the case of the window size of 48, we had to clip all parameter gradients to a maximum norm of 1 in order to prevent exploding gradients.
 	\end{itemize}
@@
 
-	Regarding to learning techniques:
+	Regarding to the learning techniques:
 
@@
 
-		\item \emph{As expected}, dropout factors applied to LSTM gates leads to far worse results than the original approach. In our case, it is not a good idea to mess with the current state of the units.
+		\item \emph{As expected}, dropout factors applied to LSTM gates lead to far worse results than the original approach.
 
@@
 
-		\item \emph{As expected}, smaller window sizes lead to worse accuracies while bigger window sizes do not offer much improvement. However, looks like the model converges in a less number of epochs.
+		\item \emph{As expected}, smaller window sizes lead to worse accuracies while bigger window sizes do not offer much improvement.
+However, it seems that the model converges in a lesser number of epochs.
 	\end{itemize}
@@
 The baseline model got a test accuracy of 0.03339 while our best model here got 0.03218---and several configurations got a mark under 0.03290---.
-We can conclude that air density, pressure and temperature help to predict wind speed at 100m in site ``90-45142''.
+We conclude that air density, pressure and temperature help to predict wind speed at 100m in site ``90-45142''.
 	\end{itemize}
@@
 
-	In this session we leverage the information of the wind speed from the other sites and we study how the MSE changes when we change the size of hidden layer, when we apply dropout and when we change the size of input window.
+	In this session we leverage the information of the wind speed from other sites and we study how the MSE is affected by the size of the hidden layer, the dropout and the size of the input window.
 
@@
 
-	Exactly the same as in section Complementary Variables.
+	Exactly the same as in Complementary Variables.
 
@@
 
-	Regarding to learning techniques: exactly the same as in section Complementary Variables.
+	Regarding to the learning techniques: exactly the same as in Complementary Variables.
 
@@
 	\begin{itemize}
-		\item \emph{Surprisingly}, we found the opposite of our expectations.
-As we can see in figure \ref{f:multistep}, with small window sizes the error seems to converge while in big window sizes the error grows exponentially.
-Sincerely, we do not have an explanation for this interesting phenomena.
+		\item \emph{Surprisingly}, as we can see in figure \ref{f:multistep}, it turns out that with small window sizes the error seems to converge while in big window sizes the error grows exponentially.
+Sincerely, we do not have a logical explanation for this interesting phenomena.
 	\end{itemize}
